---
title: "R Notebook"
output: html_notebook
---

# Gene Set Enrichment Analysis GSEA

## Define threshold for GSEA enrichments (need to be strings for cyrest call)

```{r}
fdr_DEG <- 0.05
fdr_enrichment <- 0.05

pvalue_gsea_threshold <- 0.05
qvalue_gsea_threshold <- 0.05

similarity_threshold <- 0.375
similarity_metric = "COMBINED"
#similarity_metric = "OVERLAP"
```

## Create files to be used with GSEA

```{r}
#### LIBRARIES and SOURCES ####
library(data.table)

#### SETUP ####
getwd()
setwd("C:/Users/DBC/OneDrive/Courses/Summer_2017/RNASeq")
ID_col <- "term_id"
description_col <- "term_description"
pval_col <- "pvalue"
qval_col <- "pvalue_fdr"
Genes_col <- "Genes"

dataType <- "rnaseq"
contrast <- "GSK591_Vs_SGC2096"
condition1 <- "GSK591"
condition2 <- "SGC2096"

#### STORAGE ####
folderName <- paste(dataType, contrast, sep = "_")
currentStorage <- paste(getwd(), folderName, sep = "/")
dir.create(path = currentStorage, showWarnings = TRUE)

dir.create(path = paste(currentStorage, "Fig", sep = "/"), showWarnings = TRUE)

GKPI_folders <- list("GO", "KEGG", "PFAM", "InterPro")
for (i in 1:length(GKPI_folders)) {dir.create(path = paste(currentStorage, "Fig", GKPI_folders[[i]], "Cluster_MCL", sep = "/"), showWarnings = TRUE)}

#### LOAD DATA ####
GKPI_files <- list()
for (i in 1:length(GKPI_folders)) {
  GKPI_files[[i]] <- paste(paste(GKPI_folders[[i]], dataType, contrast, "fdr5_DEGfdr5", sep = "_"), "csv", sep = ".")
}

GKPI <- list()
GKPI_filename <- list()
for (i in 1:length(GKPI_folders)) {
  GKPI[[i]] <- fread(paste(currentStorage, "Fig", GKPI_folders[[i]], GKPI_files[[i]], sep = "/"))
  
  GKPI[[i]][[Genes_col]] <- gsub(pattern = " | ", replacement = ",", x = GKPI[[i]][[Genes_col]], fixed = TRUE)
  GKPI[[i]] <- data.table(term_id = GKPI[[i]][[ID_col]], term_description = GKPI[[i]][[description_col]],
                          pvalue = GKPI[[i]][[pval_col]], pvalue_fdr = GKPI[[i]][[qval_col]],
                          phenotype = 1, Genes = GKPI[[i]][[Genes_col]])
  
  GKPI_filename[[i]] <- paste(currentStorage, "Fig", GKPI_folders[[i]], "Cluster_MCL",
                              paste(GKPI_folders[[i]], dataType, contrast, paste("fdr", qvalue_gsea_threshold*100, sep = ""), "DEGfdr5.txt", sep = "_"),
                              sep = "/")
  fwrite(GKPI[[i]], file = GKPI_filename[[i]], col.name = TRUE, sep = "\t", row.names = FALSE, quote = FALSE)
}
```

## Launch Cytoscape
Create EM through Cyrest interface - make sure you open cytoscape with a -R 1234 (to enable rest functionality) and allow R to talk directly to cytoscape.

Launch Cytoscape (by default cytoscape will automatically enable rest so as long as cytoscape 3.3 or higher is open R should be able to communicate with it). Install the App "Enrichment map" from inside Cytoscape!!

## Set up connection from R to cytoscape (already open Cytoscape)

```{r}
library(RJSONIO)

library(httr)
# Basic settings
port.number = 1234
base.url = paste("http://localhost:", toString(port.number), "/v1", sep="")

print(base.url)

version.url = paste(base.url, "version", sep="/")
cytoscape.version = GET(version.url)
cy.version = fromJSON(rawToChar(cytoscape.version$content))
print(cy.version)

enrichmentmap.url <- paste(base.url, "commands", "enrichmentmap", "build", sep="/")
```

## Create Enrichment Map

# Helper function to rename enrichment maps
```{r}
rename_cyto_network <- function(base.url, new_name, network_suid){
  
#change the name of the network
  key_value_pairs <- c()
  current <- list(id = paste(network_suid), name = new_name)
  key_value_pairs <- c(key_value_pairs, list(current))
  
  selection <- list(key = "SUID", dataKey = "id", data = key_value_pairs)
  
  selection <- toJSON(selection)
  
  update.name.url <- paste(base.url, "networks", network_suid, "tables/defaultnetwork", sep="/")
  
  response <- PUT(url = update.name.url,  
                  body = selection, encode = "json")
  return(response)
}
```

# Main

```{r}
# Get the current network ids before we add any networks.
network_ids <- fromJSON(httr::content(GET(url = paste(base.url, "networks", sep = "/")), "text", encoding = "ISO-8859-1"))

# Table with network suid to network names
network_names <- c()

# Go through each of the GSEA results files directories
for(i in 1:length(GKPI_filename)){

  cur_result <- GKPI_filename[[i]]
  cur_result_name <- paste(GKPI_folders[[i]], dataType, contrast, paste("fdr", qvalue_gsea_threshold*100, sep = ""), "DEGfdr5.txt", sep = "_")
  
  em_params <- list(analysisType = "generic",
                    pvalue = pvalue_gsea_threshold, qvalue = qvalue_gsea_threshold, similaritycutoff = similarity_threshold,
                    enrichmentsDataset1 = cur_result,
                    coeffecients = similarity_metric)

  response <- GET(url = enrichmentmap.url, query = em_params)
  
  #rename network name
  network_suid <- fromJSON(httr::content(GET(url = paste(base.url, "networks", sep = "/")), "text", encoding = "ISO-8859-1"))

  #to get the current network id figure out which is the new id
  current_network_suid <- setdiff(network_suid, network_ids)
  network_ids <-  c(network_ids, current_network_suid)
  rename_cyto_network(base.url, cur_result_name, current_network_suid)
  
  #only add if there is a network suid
  if(length(current_network_suid) > 0){
    network_names <- rbind(network_names, c(cur_result_name, unlist(current_network_suid)))
  }
}
```

## Analyze an individual network in detail 

```{r}
# Type of enrichment, i = (1 to 4)
i <- 1

EM_type <- GKPI_folders[[i]]
network_suid <- network_ids[i]

# Change EM number accordingly with the network of interest
EM <- paste("EM", i, sep = "")
```

## Cluster the Network

Find highly connected areas of the network.  Clusters contain similiar pathways that share a high percentage of genes.

```{r}
clustermaker.url <- paste(base.url, "commands", "cluster", "mcl", sep = "/") 

clustermaker_params <- list(attribute = paste(EM, "similarity_coefficient", sep = "_"))

response <- GET(url = clustermaker.url, query = clustermaker_params)

httr::content(response, as = "text", type = "html", encoding = "ISO-8859-1")
```

## Get Nodes and Edges

```{r}
getNode.url <-  paste(base.url, "networks", network_suid, "tables", "defaultnode", sep = "/") 
#returns the node table as part of data structure containing SSUID, title, public, mutable, primary key and rows
tabNode <- fromJSON(httr::content(GET(url = getNode.url), "text", encoding = "ISO-8859-1"))
default_node_table <- tabNode$rows

getEdge.url <- paste(base.url, "networks", network_suid, "tables", "defaultedge", sep = "/") 
#returns the edge table as part of data structure containing SSUID, title, public, mutable, primary key and rows
tabEdge <- fromJSON(httr::content(GET(url = getEdge.url), "text", encoding = "ISO-8859-1"))
default_edge_table <- tabEdge$rows
```

### Get the cluster information

```{r}
# Construct a Node table summary
NodeTable <- lapply(default_node_table, function(item) {
  item[[paste(EM, "Genes", sep = "_")]] <- paste(item[[paste(EM, "Genes", sep = "_")]], collapse = "|")
  item <- as.data.table(item)
})

NodeTable <- lapply(NodeTable, function(dt) {
 dt[, .SD, .SDcols = grep(pattern = paste("gs_size", "pvalue", "qvalue", "^name$", "GS_DESCR", "mclCluster", "Genes", sep = "|"),
                          x = names(dt), perl = TRUE)]
}) 

NodeTable <- rbindlist(NodeTable, fill = TRUE)

setcolorder(NodeTable, neworder = c(grep("mclCluster",names(NodeTable)), grep("gs_size",names(NodeTable)),
                                    grep("name",names(NodeTable)), grep("GS_DESCR",names(NodeTable)), grep("Genes",names(NodeTable)),
                                    grep("pvalue",names(NodeTable)),
                                    grep("fdr",names(NodeTable))
                                    ))

names(NodeTable) <- sub(pattern = paste(EM, "_", sep = ""), replacement = "", x = names(NodeTable))
setnames(NodeTable, c("ClusterNumber", "GS_Size", "Name", "GS_Description", "Genes", "pvalue", "fdr_qvalue"))

fwrite(setorderv(NodeTable, cols = "ClusterNumber", order = 1),
       file = paste(currentStorage, "Fig", EM_type, "Cluster_MCL",
                    paste(EM_type, dataType, contrast,
                          paste("fdr", qvalue_gsea_threshold*100, sep = ""),
                          paste("DEGfdr", fdr_DEG*100, sep = ""),
                          "NodeTable.csv", sep = "_"),
                    sep = "/"),
       row.names = FALSE)

# Construct an Edge table summary
EdgeTable <- lapply(default_edge_table, function(item) {
  item[[paste(EM, "Overlap_genes", sep = "_")]] <- paste(item[[paste(EM, "Overlap_genes", sep = "_")]], collapse = "|")
  item <- as.data.table(item)
})

EdgeTable <- lapply(EdgeTable, function(dt) {
  dt[, .SD, .SDcols = grep(pattern = paste("^name$", "size", "Overlap_genes", "similarity_coef", sep = "|"),
                           x = names(dt), perl = TRUE)]
})

EdgeTable <- rbindlist(EdgeTable, fill = TRUE)

setcolorder(EdgeTable, neworder = c(grep("size",names(EdgeTable)), grep("genes",names(EdgeTable)), grep("name",names(EdgeTable)),
                                    grep("similarity",names(EdgeTable))))

names(EdgeTable) <- sub(pattern = paste(EM, "_", sep = ""), replacement = "", x = names(EdgeTable))
setnames(EdgeTable, c("OverlapSize", "OverlapGenes", "Name", "SimilarityCoefficient"))

fwrite(setorderv(EdgeTable, cols = "SimilarityCoefficient", order = -1),
       file = paste(currentStorage, "Fig", EM_type, "Cluster_MCL",
                    paste(EM_type, dataType, contrast,
                          paste("fdr", qvalue_gsea_threshold*100, sep = ""),
                          paste("DEGfdr", fdr_DEG*100, sep = ""),
                          "EdgeTable.csv", sep = "_"),
                    sep = "/"),
       row.names = FALSE)
```

### Get the cluster labels 

For each cluster calculate the label of the set using Auto Annotate Cytoscape app. Using the descriptions of each geneset calculate the most frequent words in the cluster. 

```{r}
set_clusters <- unique(NodeTable$ClusterNumber)
set_clusters <- set_clusters[which(set_clusters != 0)]

ClusterTable <- list()

for(i in 1:length(set_clusters)){

    current_cluster <- set_clusters[i]
    gs_in_cluster <- NodeTable$Name[which(NodeTable$ClusterNumber == current_cluster)]
    pvalue_in_cluster <- NodeTable$pvalue[which(NodeTable$ClusterNumber == current_cluster)]
    qvalue_in_cluster <- NodeTable$fdr_qvalue[which(NodeTable$ClusterNumber == current_cluster)]
    
    # For each cluster of gs, get the gs descr to use in autoannotate
    gs_in_cluster_descr <- NodeTable$GS_Description[which(NodeTable$Name %in% gs_in_cluster)]
    descr_aa <- paste(paste(EM, "_GS_DESCR", sep = ""), gs_in_cluster_descr, sep = ":")
    
    autoannotate.url <- paste(base.url, "commands", "autoannotate", "label-clusterBoosted", sep = "/") 
    autoannotate_params <- list(labelColumn = paste(EM, "_GS_DESCR", sep = ""),
                                maxWords = 4, adjacentWordBonus = 8,
                                nodeList = paste(descr_aa, collapse = ","))

    response <- GET(url = autoannotate.url, query = autoannotate_params)
    response
    
    # Also get the Genes
    gene_in_gs_in_cluster <- NodeTable$Genes[which(NodeTable$ClusterNumber == current_cluster)]
    gene_in_gs_in_cluster <- unique(unlist(sapply(list(gene_in_gs_in_cluster), function (item) {strsplit(item, split = "|", fixed = TRUE)})))

    #something really weird happening with the AA commands.  
    #All of a sudden there are html tags attached to them
    # Haven't changed the AA version.  Not  sure why it started 
    #happening or if it is windows specific.
    
    #get rid of the html open tags
    #current_name <- strsplit(httr::content(response, "text", encoding = "ISO-8859-1"), ">")[[1]][2]
    #get rid of the html close tags
    #current_name <- strsplit(current_name, "<")[[1]][1]
    
    current_name <- gsub("\nFinished\n", "", httr::content(response, "text", encoding = "ISO-8859-1"))
    
    ClusterTable[[i]] <- data.table(ClusterNumber = current_cluster, ClusterLabel = current_name, NumberOfNodes = length(gs_in_cluster),
                                    NodesLabel = paste(gs_in_cluster, collapse = "|"), NodesDescription = paste(gs_in_cluster_descr, collapse = "|"),
                                    Genes = paste(gene_in_gs_in_cluster, collapse = "|"), NumberOfGenes = length(gene_in_gs_in_cluster),
                                    Average_pvalue = mean(pvalue_in_cluster, na.rm = TRUE), Average_qvalue = mean(qvalue_in_cluster, na.rm = TRUE)
    )
}
```

### Add the singletons as well

By default Auto annotates only adds annotations to clusters that have at least 2 nodes. Individual pathways (i.e individual nodes) are of just as much interest as clusters of them. Add annotations to the singleton nodes as well. 

```{r}
singletons <- which(is.na(NodeTable$ClusterNumber) | (NodeTable$ClusterNumber == 0))
singletonTable <- list()
if (length(singletons) > 0) { 
       for(i in 1:length(singletons)){
         name_singleton <- NodeTable$Name[singletons[i]]
         descr_singleton <- NodeTable$GS_Description[singletons[i]]
         gene_singleton <- NodeTable$Genes[singletons[i]]
         gs_size_singleton <- NodeTable$GS_Size[singletons[i]]
         pvalue_singleton <- NodeTable$pvalue[singletons[i]]
         qvalue_singleton <- NodeTable$fdr_qvalue[singletons[i]]
         singletonTable[[i]] <- data.table(ClusterNumber = paste("Singleton", i, sep = "_"), ClusterLabel = descr_singleton, NumberOfNodes = 1,
                                           NodesLabel = name_singleton, NodesDescription = descr_singleton,
                                           Genes = gene_singleton, NumberOfGenes = gs_size_singleton,
                                           Average_pvalue = pvalue_singleton, Average_qvalue = qvalue_singleton)
       }
  } else {
       message("No singleton!")
  }

# ClusterTable
ClusterTable <- rbindlist(l = list(rbindlist(ClusterTable, use.names = TRUE, fill = TRUE),
                                   rbindlist(singletonTable, use.names = TRUE, fill = TRUE)),
                          use.names = TRUE, fill = TRUE) # ALL cols are of class LIST!!

sapply(ClusterTable,class)

# Missing label name for cluster (dont know why) => check on cytoscape and manually fix it
#ClusterTable$ClusterLabel[1] <- "regulation voltage-gated calcium positive"

ClusterTable$ClusterLabel[ClusterTable$ClusterLabel == ""] <- ClusterTable$NodesDescription[ClusterTable$ClusterLabel == ""]
ClusterTable$ClusterLabel[(ClusterTable$ClusterLabel == "") & (ClusterTable$NodesDescription == "")] <- ClusterTable$NodesLabel[(ClusterTable$ClusterLabel == "") & (ClusterTable$NodesDescription == "")]


update_label <- ClusterTable[["ClusterLabel"]][grep(pattern = "|", x = ClusterTable[["ClusterLabel"]], fixed = TRUE)]
update_label <- sapply(
  strsplit(update_label, split = "|", fixed = TRUE),
  function(label) {label <- label[1]}
)
ClusterTable[["ClusterLabel"]][grep(pattern = "|", x = ClusterTable[["ClusterLabel"]], fixed = TRUE)] <- update_label

fwrite(setorderv(ClusterTable, cols = "NumberOfNodes", order = -1),
       file = paste(currentStorage, "Fig", EM_type, "Cluster_MCL",
                    paste(EM_type, dataType, contrast,
                          paste("fdr", qvalue_gsea_threshold*100, sep = ""),
                          paste("DEGfdr", fdr_DEG*100, sep = ""),
                          "ClusterTable.csv", sep = "_"),
                    sep = "/"),
       row.names = FALSE)
```

## Word cloud

```{r}
word_size <- 0.5
word_width <- 1.5e3
word_height <- 0.7e3

EM_type <- "GO"
ClusterTable <- fread(input = paste(currentStorage, "Fig", EM_type, "Cluster_MCL",
                    paste(EM_type, dataType, contrast,
                          paste("fdr", qvalue_gsea_threshold*100, sep = ""),
                          paste("DEGfdr", fdr_DEG*100, sep = ""),
                          "ClusterTable.csv", sep = "_"),
                    sep = "/"))

word <- wordcloud2(setorderv(data.table(label = ClusterTable$ClusterLabel, freq = -log(ClusterTable$Average_qvalue)), cols = "freq", order = -1), minRotation = 0, maxRotation = 0, size = word_size)
  
# Save in html
saveWidget(widget = word,
           file = paste(currentStorage, "Fig", EM_type, "Cluster_MCL",
                        paste(
                          paste(EM_type, "WordCluster", dataType, contrast,
                                paste("fdr", fdr_enrichment*100, sep = ""),
                                paste("DEGfdr", fdr_DEG*100, sep = ""),
                                sep = "_"),
                          "html",
                          sep = "."),
                        sep = "/"), 
           selfcontained = F)
  
# Save in png
webshot::webshot(url = paste(currentStorage, "Fig", EM_type, "Cluster_MCL",
                             paste(
                               paste(EM_type, "WordCluster", dataType, contrast,
                                     paste("fdr", fdr_enrichment*100, sep = ""),
                                     paste("DEGfdr", fdr_DEG*100, sep = ""),
                                     sep = "_"),
                               "html",
                               sep = "."),
                             sep = "/"),
                 file = paste(currentStorage, "Fig", EM_type, "Cluster_MCL",
                              paste(
                                paste(EM_type, "WordCluster", dataType, contrast,
                                      paste("fdr", fdr_enrichment*100, sep = ""),
                                      paste("DEGfdr", fdr_DEG*100, sep = ""),
                                      sep = "_"),
                                "png",
                                sep = "."),
                              sep = "/"),
                 delay = 20, vwidth = word_width, vheight = word_height)

```

